---
title: "Agglomerative Hierarchical Clustering"
description: |
  This code will use hierarchical clustering by complete linkage to create a dendrogram showing multivariate clustering for water chemistry by site.
author:
  - name: Taylor Gries
    url: {}
date: 2022-03-01
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide 
---


# Overview

This code will use hierarchical clustering by complete linkage to create a dendrogram showing multivariate clustering for water chemistry by site. This data looks at Santa Barbara Coastal drainage are and has been ongoing since 2000. 

## Variables

- Total Dissolved Nitrogen uM
- Total Dissolved Phosphorus uM
- Total Particulate Carbon micromolesPerLiter
- Total Particulate Nitrogen micromolesPerLiter
- Total Suspended Solids (milligramPerLiter)


**Data Citation:** Santa Barbara Coastal LTER and J. Melack. 2019. SBC LTER: Land: Stream chemistry in the Santa Barbara Coastal drainage area, ongoing since 2000 ver 16. Environmental Data Initiative. 

```{r setup, include=TRUE, message =FALSE, warning = FALSE, echo = TRUE, error= TRUE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(janitor)
library(here)
library(naniar)
 
# Packages for cluster analysis:
library(NbClust)
library(cluster)
library(factoextra)
library(dendextend)
library(ggdendro)
```


```{r}
stream_chem <- read_csv( "sbc_lter_registered_stream_chemistry.csv")
```

# Data Wrangling

To perform hierarchical clustering by site, I made a data frame that has a single summary row per site (e.g. based on means from all observations at that site). Then, I calculated the Euclidean distance before performing complete linkage agglomerative hierarchical clustering. 

```{r}
stream_chem_clean <- stream_chem %>%
  na_if(-999) %>%
  clean_names() %>%
  group_by(site_code) %>%
  select(site_code, nh4_u_m:spec_cond_u_spercm) %>%
  summarize(
    "Ammonium" = mean(nh4_u_m, na.rm = TRUE),
    "Nitrate" = mean(no3_u_m, na.rm = TRUE),
    "Phosphorous" = mean(po4_u_m, na.rm = TRUE),
    "Dissolved Nitrogen" = mean(tdn_u_m, na.rm = TRUE),
    "Dissolved Phosphorous" = mean(tdp_u_m, na.rm = TRUE),
    "Particulate Carbon" = mean(tpc_u_m, na.rm = TRUE),
    "Particulate Nitrogen" = mean(tpn_u_m, na.rm = TRUE),
    "Particulate Phosphorous" = mean(tpp_u_m, na.rm = TRUE),
    "Suspended Solids" = mean(tss_mgper_liter, na.rm = TRUE),
    "Specific Conductivity" = mean(spec_cond_u_spercm, na.rm = TRUE))
```


```{r}
stream_chem_scale <- stream_chem_clean %>% 
  select(-site_code) %>%
  scale() # See ?scale for details on scaling

# compare scaled to original vars
summary(stream_chem_scale)
```

# Dendrogram 

## Complete Linkage

```{r}
rownames(stream_chem_scale) <- stream_chem_clean$site_code # adding rownames from original dataset
```

**Figure 1**: A cluster dendrogram that illustrates water chemistry by site using complete linkage - the clusters are merged by the smallest **maximum** distance between two observations in distinct clusters. *Data: Santa Barbara Coastal LTER and J. Melack. (2019)*


```{r}
stream_chem_dist <- dist(stream_chem_scale, method = 'euclidean') 

# Hierarchical clustering (complete linkage)
stream_hc_complete <- hclust(stream_chem_dist, method = "complete")


# Pretty plot
ggdendrogram(stream_hc_complete, 
             rotate = TRUE) +
  theme_minimal() +
  labs(title = "Water Chemistry by Site (Complete Linkage)",
       x = "Stream Site",
       y = "Height",
       caption = "Data: Santa Barbara Coastal LTER and J. Melack (2019).")

```


## Single Linkage

**Figure 2**: Cluster dendrogram for illustrating water chemistry by site using single linkage - the clusters are merged by the *smallest* distance between observations in separate clusters. *Data: Santa Barbara Coastal LTER and J. Melack. (2019)*


```{r}
# Hierarchical clustering (single linkage)
stream_hc_single <- hclust(stream_chem_dist, method = "single")


# Pretty plot
ggdendrogram(stream_hc_single, 
             rotate = TRUE) +
  theme_minimal() +
  labs(title = "Water Chemistry by Site (Single Linkage)",
       x = "Stream Site",
       y = "Height",
       caption = "Data: Santa Barbara Coastal LTER and J. Melack (2019).")

```


# Results

- There is a large mean concentration of Ammonium in stream 4 (DV01) which is the largest grouping

- There are 3 sites in one of these two sub-groupings that all have high Ammonium concentrations compared to other sites. 

- Since 9 out of 13 sites fit into one sub-grouping, this could mean the variables are correlated in some way for most of the sites.

- All other sites are included in a grouping with at least two sub-groupings.

- Further investigation is needed for streams not included in the larger sub-groups. 

# Citations

**Data Citation:** Santa Barbara Coastal LTER and J. Melack. 2019. SBC LTER: Land: Stream chemistry in the Santa Barbara Coastal drainage area, ongoing since 2000 ver 16. Environmental Data Initiative. 


