---
title: "Data Visualization: Mono Lake Water Levels from 1850 - 2017"
description: |
  This code wrangles data relating to Mono Lake water levels from 1850 to 2017.
author:
  - name: Taylor Gries
    url: {}
date: 2022-03-01
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---



```{r setup, message = FALSE, warning = FALSE, echo = TRUE}
# Load in packages
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(janitor)
library(GGally)
library(kableExtra)
library(modelsummary)
library(stargazer)
library(effsize)
```

```{r, include = FALSE}
mono_lake <- read_csv("mono_lake_annual_levels.csv", skip = 5)
```


```{r, include=FALSE}
mono_lake_clean <- mono_lake %>%
  clean_names() # Clean column names
```


```{r}
ggplot(data = mono_lake_clean, aes(x = year, y = lake_level_feet_above_sea_level)) + # Create line graph
  geom_line(color = "darkblue", 
            size = 0.4) + # Change color and size of lake level data
  scale_x_continuous(limits = c(1850, 2020),
                     expand = c(0,0),
                     breaks = seq(1850, 2020, by = 25)) + # Fix x grid
  scale_y_continuous(limits = c(6372, 6428),
                     expand = c(0,0),
                     breaks = seq(6370, 6430, by = 10)) + # Fix y grid
  theme_minimal() + # Change theme 
  labs(x = "Year", # Add in plot labels
       y = "Water Level (Ft above Sea Level)",
       title = "Annual Mono Lake Water Levels: 1850 - 2017",
       caption = "Data: Mono Basin Clearinghouse") +
  theme(axis.title.x = element_text(size = 10, # Edit text 
                                    hjust = 0.5),
        axis.title.y = element_text(size = 10, 
                                    hjust = 0.5),
        plot.title = element_text(size = 14, 
                                  hjust = 0.5)) +
    geom_hline(yintercept = 6392, # Add in stable water level line
               linetype = "solid", 
               color = "darkgreen",
               size = 0.5) +
  geom_text(x = 1852, y = 6390, # Add in stable water level text
            label = "Stable Lake Level at 6,392 ft",
            color = "darkgreen",
            size = 3,
            hjust = 0,
            vjust = 0.5) +
  geom_hline(yintercept = 6377, # Add in Land Bridge line
             linetype = "dashed",
             color = "darkred",
             size = 0.5) +
  geom_text(x = 1852, y = 6375, 
            label = "Land Bridges Emerge at 6,377 ft", # Add in Land Bridge text
            color = "darkred",
            size = 3,
            hjust = 0,
            vjust = 0.5) +
  annotate("text", label = "1941: LA DWP Begins Diverting Water", # Add in historical event text
           x = 1950,
           y = 6420,
           size = 2.5,
           vjust = 0,
           hjust = 0.15) +
  geom_point(aes(x = 1941, y = 6417), # Add in point for historical event
             size = 1) +
  annotate("text", label = "1976: First Ecological Study", # Add in historical event text
           x = 1965,
           y = 6380,
           size = 2.5,
           vjust = .1,
           hjust = .8) +
  geom_point(aes(x = 1976, y = 6378), # Add in point for historical event
             size = 1) +
  annotate("text", label = "1998: State Issued Restoration Orders", # Add in historical event text
           x = 1995,
           y = 6388,
           size = 2.5,
           vjust = .2,
           hjust = .45) +
  geom_point(aes(x = 1998, y = 6384.4), # Add in point for historical event
             size = 1)
```

**Figure 1:** Mono Lake levels have plumented since 1941 in a 40 year time span due to the Los Angeles Department of Water & Power (DWP) diverting water, causing the lake to los half its volume and doubling in salinity. Once students discovered the ecological collapse in 1976, action has been taken to decrease land bridges causing a decline in bird nesting sites and to increase food resources for aquatic birds. 




